Diversity dynamics in Nymphalidae butterflies: Effect of phylogenetic uncertainty on diversification rate shift estimates

### 2015-01-14 MS_orig.tex
ready for resubmittion
revision 2

\textbf{Comments:} 5814 words, 7 figures, 2 tables and 25 supplementary
material files.

\textbf{Additional Supplementary materials:}
\url{http://dx.doi.org/10.6084/m9.figshare.639208}
\end{flushleft}

@@

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects {[}1{]}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects {[}2{]}. The
``escape-and-radiate'' hypothesis {[}3{]}, the ``oscillation
hypothesis'' {[}4,5{]} or ``diffuse cospeciation'' {[}2{]}.

The butterfly family Nymphalidae has been an important taxon for
developing some of the mentioned hypotheses. Nymphalidae contains around
6000 species {[}6{]}, and is the largest family within the true
butterflies. The family most likely originated around 94 MYA in the mid
Cretaceous. Diversification of the group began in the Late Cretaceous
and most major radiations (current subfamilies) appeared shortly after
the Cretaceous-Paleogene (K-Pg) boundary {[}7{]}. Several studies have
used time-calibrated phylogenies and diversification models to
reconstruct the evolutionary history of the group to identify patterns
of accelerated or decelerated diversification of some Nymphalidae clades
{[}7--10{]}. For example, it has been suggested that climate change in
the Oligocene and the subsequent diversification of grasses has led to
diversification of the subfamily Satyrinae {[}11{]} due to the abundance
of grasses over extensive geographic areas (``resource
abundance-dependent diversity dynamics'' hypothesis). Reference 9 found
increased net diversification rates in some Nymphalidae lineages after a
major hostplant shift, which appears to be in agreement with Reference 3
``escape-and-radiate'' model of diversification.

Although it has been suggested that part of the great diversity of
Nymphalidae butterflies is a result of hostplant-insect dynamics, it is
necessary to use modern techniques to investigate whether the
diversification patterns of Nymphalidae are in agreement with the
theoretical predictions. It is necessary to test whether the overall
diversification pattern of Nymphalidae is congruent with events of
sudden diversification bursts due to hostplant shift, climatic events or
shifts to closely related hostplants {[}5,12{]}.

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies {[}10{]} to investigate patterns
of diversification. We applied MEDUSA {[}13,14{]}, a recently developed
statistical method, to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA {[}13{]}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} {[}15{]}.

\section*{Methods}

@@

For analyses, we used the phylogenetic trees from the study of Reference
10 that were generated using DNA sequence data from 10 gene regions for
398 of the 540 valid genera in Nymphalidae. We employed Reference 10
maximum clade credibility tree (MCC tree) (Fig. 1) as well as a random
sample of 1000 trees from their BEAST run after burnin. Their original
BEAST run was for 40 million generations. We used a burnin of 25 million
generations and took a random sample of 1000 trees using Burntrees
v.0.1.9 \url{http://www.abc.se/~nylander/} (supp. mat. 5) in order to
correct for phylogenetic uncertainty when performing the diversification
analyses.

Species richness data for Nymphalidae genera were compiled from several
sources including the specialist-curated lists on
\url{http://tolweb.org}, Reference 16 and curated lists of Global
Butterfly Names project (\url{http://www.ucl.ac.uk/taxome/gbn/}). We
assigned the species numbers of genera not included in the phylogeny to
the closest related genus that was included in Reference 10 study
according to available phylogenetic studies {[}17--28{]}, taxonomical
classification and morphological resemblance when no phylogenies were
available (supp. mat. 3).

Hostplant data for Nymphalidae species were compiled from several
sources including Reference 29, HOSTS database
(\url{http://bit.ly/YI7nwW}), Reference 30, Reference 31, Reference 32
and others (supp. mat. 15--16) for a total of 6586 hostplant records,
including 428 Nymphalidae genera and 143 plant families and 1070 plant
genera. It was not possible to find any hostplant data for 35 butterfly
genera.

@@

We used the statistical software R version 3.0.1 {[}33{]} in combination
with the APE {[}34{]}, GEIGER {[}35{]}, MEDUSA {[}13{]} and
\texttt{diversitree} {[}15{]} packages along with our own scripts to
perform the analyses (included as supplementary materials 2, 6, 10). All
analyses were run on the 1000 random trees from Reference 10, on the
maximum clade credibility tree derived from these and the MCC tree from
Reference 10.

@@

Patterns of diversification in Nymphalidae were analyzed by using MEDUSA
version 093.4.33 {[}13{]} on the maximum clade credibility tree from
Reference 10. MEDUSA fits alternative birth-death likelihood models to a
phylogenetic tree in order to estimate changes in net diversification
rates along branches. MEDUSA estimates likelihood and AIC scores for the
simplest birth-death model, with two parameters, the rates \texttt{r}:
net diversification \texttt{=} speciation (\texttt{b}) \texttt{-}
extinction (\texttt{d}) and \(\varepsilon\): relative extinction
\texttt{= d/b}. The AIC scores of the two-parameter model are then
compared with incrementally more complex models until the addition of
parameters do not improve the AIC scores beyond a cutoff value. MEDUSA
finds the likelihood of the models after taking into account branch
lengths and number of species per lineage {[}13{]}. Most studies using
MEDUSA only run the method on a single tree, usually the maximum clade
credibility tree, which makes the assumption that this tree is correct.
We wanted to study the effects of phylogenetic uncertainty on estimation
of net diversification rate shifts and therefore selected 1000 random
genus-level trees from the posterior distribution (MultiMEDUSA, supp.
mat. 5). We calculated a new MCC tree from the selection of trees, ran
MEDUSA and MultiMEDUSA on the selection of 1000 trees and summarized the
estimated changes in net diversification rates for nodes across all
trees. Patterns of change in net diversification rates are considered
significant if they are found at the same node in at least 90\% of the
trees. We also expected to find similar values of net diversification
(\texttt{r = b - d}) and relative extinction rate (\(\varepsilon\)
\texttt{= d/b}) values across the 1000 trees for the nodes where changes
in diversification tempo occurs. We used the AICc threshold of 7.8
units, as estimated by MEDUSA, as the limit for a significantly better
fit to select among increasingly complex alternative models.

@@

As the MEDUSA and MultiMEDUSA approaches estimated an increase in net
diversification in the clade Ithomiini, we tested whether this pattern
can be explained by increase in birth-rate due to hostplant use and
performed analyses with the ``binary state speciation and extinction''
{[}36{]} Bayesian approach as implemented in the R package
\texttt{diversitree} {[}15{]}. MuSSE {[}15{]} is designed to examine the
joint effects of two or more traits on speciation. Because most of
Nymphalidae butterflies are restricted to use one plant family as
hostplant, the character states can be coded as presence/absence, for
which the BiSSE analysis is better suited. BiSSE was designed to test
whether a binary character state has had any effect on increased net
diversification rate for a clade {[}36{]}. We used our compiled data of
hostplant use to produce a binary dataset for the character ``feeding on
the plant family Solanaceae'' with two states (absence = 0; presence =
1) (supp. mat. 17), which is the main hostplants of the diverse
Ithomiini butterflies and closest relatives {[}37{]}. Other hostplant
shifts were not tested for effect on net diversification rates due to
the low support for diversification shifts found in the MEDUSA analyses.
We analyzed the data using BiSSE employing the Markov Chain Monte Carlo
algorithm on the maximum clade credibility tree, taking into account
missing taxa by using the parameter ``sampling factor''
(\texttt{sampling.f}) in \texttt{diversitree}. We also used constrained
analyses forcing no effect of hostplant use on diversification and used
likelihood ratio tests to find out whether the hypothesis of effect on
diversification has a significantly better likelihood than the null
hypothesis (no effect). The analyses were run across a sample of 250
trees from the posterior distribution. The records of \emph{Vanessa} and
\emph{Hypanartia} feeding on Solanaceae {[}30,38{]} might be incorrect
as it is unlikely that these species can be feeding on this plant
family. We tested whether coding these two genera as not feeding on
Solanaceae affected our results.

@@

The MEDUSA analysis on the MCC tree in combination with richness data
estimated 18 changes in the tempo of diversification in Nymphalidae
history (Fig. 1; Table 1). The corrected AICc acceptance threshold for
adding subsequent piecewise birth-death processes to the overall model
was set to 7.8 units, as prescribed by MEDUSA. In all MEDUSA analyses,
the maximum number of inferred diversification splits in all trees was
26. The background net diversification rate for Nymphalidae was
estimated as \texttt{r = 0.092} lineages per Million of years and the
AICc score for the best fit model was \texttt{5090.5} (Table 1). MEDUSA
also estimated that the basic constant birth--death model was not a
better explanation for our data (AICc \texttt{= 5449.3}).

Some of the 18 changes in diversification correspond to rate increases
in very species-rich genera: \emph{Ypthima} (\texttt{r = 0.311}),
\emph{Charaxes} (\texttt{r = 0.291}), \emph{Callicore} +
\emph{Diaethria} (\texttt{r = 0.220}), \emph{Pedaliodes}
(\texttt{r = 0.196}) and \emph{Taenaris} (\texttt{r = 0.234}). We found
rate increases for other clades as well such as: Mycalesina
(\texttt{r = 0.191}), Oleriina + Ithomiina + Napeogenina + Dircennina +
Godyridina (\texttt{r = 0.187}), Satyrini (\texttt{r = 0.116}),
Phyciodina in part (\texttt{r = 0.241}) and Satyrina
(\texttt{r = 0.221}), \emph{Coenonympha} (\texttt{r = 0.209}),
\emph{Caeruleuptychia} + \emph{Magneuptychia} (\texttt{r = 0.312}) and
\emph{Taenaris} (\texttt{r = 0.312}). We also found decreases in net
diversification rates for Limenitidinae + Heliconiinae
(\texttt{r = 0.0541}), part of Danaini (\texttt{r = 0.0423}),
Pseudergolinae (\texttt{r = 0.024}) and Coenonymphina
(\texttt{r = 0.065}) (Table 1).

@@

We tested MEDUSA to find out whether taking into account the
phylogenetic signal from a random sample of 1000 trees from the
posterior distribution can return similar estimates of diversification
to the values obtained from the MCC tree. We run MultiMEDUSA on a random
sample of 1000 trees (supp. mat. 5) from the posterior distribution and
compared the results with a MEDUSA analysis on the MCC tree derived from
this sample (supp. mat. 9).

We found that the analysis by MultiMEDUSA on the 1000 trees estimated
lower median net diversification rates for the diversification shifts
found by MEDUSA on the MCC tree from Reference 10 (Table 2). Although
the diversification pattern found by MEDUSA and MultiMEDUSA was the
same, the latter consistently estimated lower rates. Thus, the shifts
recovered with low net diversification rate on the MCC were recovered
with negative net diversification rate by MultiMEDUSA. The background
diversification and all shifts found by MEDUSA on the 1000 trees are
provided as an R object in supp. mat. 7).

We also compared the results from MultiMEDUSA (derived from the sample
of 1000 trees) with the splits found by MEDUSA on the MCC tree derived
from this random sample. In the summary statistics, MultiMEDUSA reports
the frequency of the diversification shifts found in the trees
(parameter \texttt{sum.prop}). Thus, if a node is found in only half of
the 1000 trees, but the phylogenetic signal was strong enough to be
picked up by MEDUSA and a node shift was found most of the time, then
the \texttt{sum.prop} should be close to 1. For example the node 625
consists of the clade \texttt{Charaxes} + \texttt{Polyura} and it was
found in only 256 trees, however MultiMEDUSA was consistenly able to
find a diversification shift for that node and the \texttt{sum.prop}
value is 0.996.

We found different rates for diversification shifts even for the nodes
that appear in most of the sample of 1000 trees (Table 2) as many of the
nodes also found in the MCC tree had a probability of being recovered
higher than 0.90.

There were four diversification shifts found with more than 90\% of
probability in the trees from the random sample (Table 2): (i) net
diversification rate increase in the genus \emph{Ypthima}; (ii) net
diversification rate increase in the genus \emph{Charaxes}; (iii) rate
increase in Ithomiini subtribes Oleriina + Ithomiina + Napeogenina +
Dircennina + Godyridina; and (iv) rate increase in \emph{Callicore} +
\emph{Diaethria}.

MultiMEDUSA provided mean and standard deviation statistics for the
diversification values found on the shifts on the 1000 trees (supp. mat.
13--14), and found that some of the changes in net diversification rate
values had great variation across the posterior distribution of trees. A
boxplot of the net diversification rate values estimated for the clades
that appear in the MCC tree shows that some shifts are estimated as
increased or slowed diversification pace depending on the tree used for
analysis (Fig. 2). This variation is especially wide for the clade
formed by the genera \emph{Magneuptychia} and \emph{Caeruleuptychia}
because MEDUSA estimated diversification values from six times the
background net diversification rate (\texttt{r = 0.5234}) to almost zero
(\texttt{r = 1.22e-01}). The rates for \emph{Taenaris} were between 0.14
and 0.44 (mean value 0.25). Similar degrees of variation were found in
the nodes for \emph{Ypthima}, \emph{Charaxes} and \emph{Coenonympha}
(Fig. 2). The net diversification rates estimates for the clades
(Oleriina + Ithomiina + Napeogenina + Dircennina + Godyridina),
Limenitidinae + Heliconiinae and Pseudergolinae are relatively
consistent across the 1000 trees (Fig. 2).

It is also evident that not all the diversification shifts estimated on
the MCC tree are consistently recovered in most of the 1000 trees. Some
of the splits in the MCC tree are recovered in very few trees, for
example the split for the clade Satyrini (Euptychiina + Pronophilina +
Satyrina + Maniolina) is recovered with a probability of 18\% (Fig. 3).

@@

The MEDUSA analyses taking into account phylogenetic uncertainty
estimated a net diversification rate increase in part of the clade
Ithomiini across more than 95\% of the trees. Our BiSSE analysis found a
positive effect of the character state ``feeding on Solanaceae'' on the
net net diversification rate on part of Ithomiini (Oleriina + Ithomiina
+ Napeogenina + Dircennina + Godyridina) (Fig. 4). The Markov Chain
Monte Carlo algorithm was run for 10000 generations discarding the first
7500 as burnin. The estimated mean net diversification rate for taxa
that do not feed on Solanaceae was \texttt{r = 0.11} while the net
diversification rate for the Solanaceae feeders was \texttt{r = 0.16}
(see Fig. S1 for a boxplot of speciation and extinction values for the
95\% credibility intervals). The same analysis considering
\emph{Vanessa} and \emph{Hypanartia} as non-Solanaceae feeders due to
dubious records produced the same pattern and net diversification rates
(Fig. S2). Therefore, the rest of the analyses were performed assuming
these two genera as Solanaceae feeders. We constrained the BiSSE
likelihood model to force equal rates of speciation for both character
states in order to test whether the model of different speciation rates
is a significantly better explanation for the data. A likelihood ratio
test found that the model for increased net diversification rate for
nymphalids feeding on Solanaceae is a significantly better explanation
than this character state having no effect on diversification
(\texttt{p \textless{} 0.001}) (Table 3; character states available in
supp. mat. 17, code in supp. mat. 18, and mcmc run in supp. mat. 19). We
combined the post-burnin mcmc generations from running BiSSE on 250
trees from the posterior distribution and found the same pattern as the
BiSSE analysis on the maximum clade credibility tree (combined mcmc run
in supp. mat. 20; profiles plot of speciation rates in Fig. S3; boxplot
of 95\% credibility intervals in Fig. S4). A BiSSE analysis to test
whether the trait ``feeding on Apocynaceae'' had any effect on increased
net diversification rates found similar speciation rates for lineages
feeding on Apocynaceae and other plants (Fig. S5). It has been shown
that BiSSE performs poorly under certain conditions {[}39{]}. However,
our data has adequate number of taxa under analysis (more than 300
tips), adequate speciation bias (between 1.5x and 2.0x), character state
bias (around 8x) and extinction bias (around 4x) for the analysis of
Solanaceae hostplants. Thus, BiSSE is expected to produce robust results
{[}39{]}.

@@

\paragraph*{Ithomiini}

>>>>

\end{flushleft}

@@

Hostplant shifts have been invoked to be responsible for a great part of
the biodiversity of herbivorous insects \cite{mitter1988}. The study of the
evolution of hostplant use has spawned several theories explaining the
evolutionary interactions between plants and insects \cite{nyman2012}. The
``escape-and-radiate'' hypothesis \cite{ehrlich1964}, the ``oscillation
hypothesis'' \cite{janz2011, nylin2014} or ``diffuse cospeciation'' \cite{nyman2012} and the
``musical chairs hypothesis'' \cite{hardy2014}.

The butterfly family Nymphalidae has been an important taxon for
developing some of the mentioned hypotheses. Nymphalidae contains around
6000 species \cite{van_nieukerken2011}, and several members are considered model organisms
in evolutionary biology \cite{joron2006, willmott2006, brakefield2009}.
The family most likely originated
around 94 MYA in the mid Cretaceous. Diversification of the group began
in the Late Cretaceous and most major radiations (current subfamilies)
appeared shortly after the Cretaceous-Paleogene (K-Pg) boundary
\cite{heikkila2012}. Several studies have used time-calibrated phylogenies and
diversification models to reconstruct the evolutionary history of the
group to identify patterns of accelerated or decelerated diversification
of some Nymphalidae clades \cite{heikkila2012, elias2009, fordyce2010, wahlberg2009}.
For example, it has been
suggested that climate change in the Oligocene and the subsequent
diversification of grasses has led to diversification of the subfamily
Satyrinae \cite{pena2008} due to the abundance of grasses over extensive
geographic areas (``resource abundance-dependent diversity dynamics''
hypothesis): Fordyce (2010) \cite{fordyce2010} found increased net diversification
rates in some Nymphalidae lineages after a major hostplant shift, which
appears to be in agreement with the ``escape-and-radiate'' model of
diversification \cite{ehrlich1964}.

Although it has been suggested that part of the great diversity of
Nymphalidae butterflies is a result of hostplant-insect dynamics, it is
necessary to use modern techniques to investigate whether the
diversification patterns of Nymphalidae are in agreement with the
theoretical predictions. It is necessary to test whether the overall
diversification pattern of Nymphalidae is congruent with events of
sudden diversification bursts due to hostplant shift or climatic events
\cite{nylin2014, ferrer2013}.

In this study, we used a time-calibrated genus-level phylogenetic
hypothesis for Nymphalidae butterflies \cite{wahlberg2009} to investigate patterns
of diversification. We applied the statistical method MEDUSA
\cite{alfaro2009, harmon2011}, to study the diversification pattern of Nymphalidae
butterflies. MEDUSA fits likelihood models of diversification into a
time-calibrated tree and tests whether allowing increases or decreases
in speciation and extinction rates within the tree produces better fit
of the models. MEDUSA is able to take into account unsampled extant
species diversity during model fitting and it is normally applied to the
maximum clade credibility phylogenetic tree. Particularly, we wanted to
study the effects of phylogenetic uncertainty and by using the extended
MEDUSA method called MultiMEDUSA \cite{alfaro2009}. We also tested whether
hostplant association dynamics can explain the diversification patterns
of component Nymphalidae lineages by testing whether character states of
hostplant use affected the diversification pattern of those lineages
employing the method BiSSE as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}.

\section*{Materials and Methods}

@@

For analyses, we used the phylogenetic trees from the study of Wahlberg
et al. (2009) \cite{wahlberg2009} that were generated using DNA sequence data from
10 gene regions for 398 of the 540 valid genera in Nymphalidae. We
employed the maximum clade credibility tree (MCC tree) (Fig. 1) as well
as a random sample of 1000 trees from their BEAST run after burnin
\cite{wahlberg2009}. Their original BEAST run was for 40 million generations. We
used a burnin of 25 million generations and took a random sample of 1000
trees using Burntrees v.0.1.9 \url{http://www.abc.se/~nylander/} (supp.
mat. 5) in order to correct for phylogenetic uncertainty when performing
the diversification analyses.

Species richness data for Nymphalidae genera were compiled from several
sources including the specialist-curated lists on
\url{http://tolweb.org}, Lamas (2004) \cite{lamas2004} and curated lists of
Global Butterfly Names project (\url{http://www.ucl.ac.uk/taxome/gbn/}).
We assigned the species numbers of genera not included in the phylogeny
to the closest related genus that was included in Wahlberg et al. (2009)
\cite{wahlberg2009} study according to available phylogenetic studies 
\cite{matos2013, brower2010, kodandaramaiah2010, kodandaramaiah2010a,
ortiz2013, desilva2010, freitas2004, pena2006, penz1999, silva2008,
pena2011, pena2010},
taxonomical classification and morphological resemblance when no
phylogenies were available (supp. mat. 3).

Hostplant data for Nymphalidae species were compiled from several
sources including \cite{ackery1988}, HOSTS database (\url{http://bit.ly/YI7nwW}),
\cite{dyer2002, beccaloni2008, janzen2009} and others (supp. mat. 15--16) for a total of 6586
hostplant records, including 428 Nymphalidae genera and 143 plant
families and 1070 plant genera. It was not possible to find any
hostplant data for 35 butterfly genera.

@@

We used the statistical software R version 3.0.1 \cite{r2013} in combination
with the APE \cite{popescu2012}, GEIGER \cite{harmon2008}, MEDUSA
\cite{alfaro2009} and
\texttt{diversitree} \cite{fitzjohn2012} packages along with our own scripts to
perform the analyses (included as supplementary materials 2, 6, 10). All
analyses were run on the 1000 random trees from Wahlberg et al. (2009)
\cite{wahlberg2009}, on the maximum clade credibility tree derived from these and
the MCC tree \cite{wahlberg2009}.

We analyzed our MCC tree in the Bayesian software BAMM in order to
contrast the results found by MEDUSA and MultiMEDUSA. BAMM (Bayesian
analysis of macroevolutionary mixtures) \cite{rabosky2014} is a software that fits
several models of speciation and extinction. The best model can be
selected by comparison of Bayes Factors.

@@

Patterns of diversification in Nymphalidae were analyzed by using MEDUSA
version 093.4.33 \cite{alfaro2009} on the maximum clade credibility tree from
Wahlberg et al. (2009) \cite{wahlberg2009}. MEDUSA fits alternative birth-death
likelihood models to a phylogenetic tree in order to estimate changes in
net diversification rates along branches. MEDUSA estimates likelihood
and AIC scores for the simplest birth-death model, with two parameters,
the rates \texttt{r}: net diversification \texttt{=} speciation
(\texttt{b}) \texttt{-} extinction (\texttt{d}) and (\(\varepsilon\):
relative extinction \texttt{= d/b}. The AIC scores of the two-parameter
model are then compared with incrementally more complex models until the
addition of parameters do not improve the AIC scores beyond a cutoff
value. MEDUSA finds the likelihood of the models after taking into
account branch lengths and number of species per lineage \cite{alfaro2009}. Most
studies using MEDUSA only run the method on a single tree, usually the
maximum clade credibility tree, which makes the assumption that this
tree is correct. We wanted to study the effects of phylogenetic
uncertainty on estimation of net diversification rate shifts and
therefore selected 1000 random genus-level trees from the posterior
distribution of the Bayesian run from Wahlberg et al., 2009 \cite{wahlberg2009}
(MultiMEDUSA, supp. mat. 5). We calculated a new MCC tree from the
selection of trees, ran MEDUSA and MultiMEDUSA on the selection of 1000
trees and summarized the estimated changes in net diversification rates
for nodes across all trees. Patterns of change in net diversification
rates are considered significant if they are found at the same node in
at least 90\% of the trees. We also expected to find similar values of
net diversification (\texttt{r = b - d}) and relative extinction rate
(\(\varepsilon\) \texttt{= d/b}) values across the 1000 trees for the
nodes where changes in diversification tempo occurs. We used the AICc
threshold of 7.8 units, as estimated by MEDUSA, as the limit for a
significantly better fit to select among increasingly complex
alternative models.

@@

As the MEDUSA and MultiMEDUSA approaches estimated an increase in net
diversification in the clade Ithomiini, we tested whether this pattern
can be explained by an increase in the birth-rate due to hostplant use
and performed analyses using the ``binary state speciation and
extinction'' \cite{maddison2007} Bayesian approach as implemented in the R package
\texttt{diversitree} \cite{fitzjohn2012}. The Markov Chain Monte Carlo algorithm
was run for 10000 generations discarding the first 7500 as burnin. The
Multiple State Speciation and Extinction approach (MuSSE) \cite{fitzjohn2012} is
designed to examine the joint effects of two or more traits on
speciation. Because most of Nymphalidae butterflies are restricted to
use one plant family as hostplant (232 genera use only one plant family
versus 176 that use more than one; based on our data in supp. mat. 15),
the character states can be coded as presence/absence, for which the
BiSSE analysis is better suited. BiSSE was designed to test whether a
binary character state has had any effect on increased net
diversification rate for a clade \cite{maddison2007}. We used our compiled data of
hostplant use to produce a binary dataset for the character ``feeding on
the plant family Solanaceae'' with two states (absence = 0; presence =
1) (supp. mat. 17). The family Solanaceae include the hostplants of the
diverse Ithomiini butterflies and closest relatives \cite{willmott2006}. Other
hostplant shifts were not tested for effect on net diversification rates
due to the low support for diversification shifts found in the MEDUSA
analyses. We analyzed the data using BiSSE employing the Markov Chain
Monte Carlo algorithm on the maximum clade credibility tree, taking into
account missing taxa by using the parameter ``sampling factor''
(\texttt{sampling.f}) in \texttt{diversitree}. We also used constrained
analyses forcing no effect of hostplant use on diversification and used
likelihood ratio tests to find out whether the hypothesis of effect on
diversification has a significantly better likelihood than the null
hypothesis (no effect). The analyses were run across a sample of 250
trees from the selected 1000 random trees from the posterior
distribution. The records of \emph{Vanessa} and \emph{Hypanartia}
feeding on Solanaceae \cite{beccaloni2008, scott1986} might be incorrect as it is unlikely
that these species can be feeding on this plant family. We tested
whether coding these two genera as not feeding on Solanaceae affected
our results. We expected that coding these two genera as either absence
or presence would not have a significant effect on our results. It has
been shown that BiSSE performs poorly under certain conditions \cite{davis2013}.
However, our data has adequate number of taxa under analysis (more than
300 tips), adequate speciation bias (between 1.5x and 2.0x), character
state bias (around 8x) and extinction bias (around 4x) for the analysis
of Solanaceae hostplants. Thus, BiSSE is expected to produce robust
results \cite{davis2013}.

@@

The MEDUSA analysis on the MCC tree in combination with richness data
estimated 18 significant changes in the tempo of diversification in
Nymphalidae history (Fig. 1; Table 1). The corrected AICc acceptance
threshold for adding subsequent piecewise birth-death processes to the
overall model was set to 7.8 units, as prescribed by MEDUSA. In all
MEDUSA analyses, the maximum number of inferred diversification splits
in all trees was 26. The background net diversification rate for
Nymphalidae was estimated as \texttt{r = 0.092} lineages per Million of
years and the AICc score for the best fit model was \texttt{5090.5}
(Table 1). MEDUSA also estimated that the basic constant birth-death
model was not a better explanation for our data (AICc
\texttt{= 5449.3}).

Some of the 18 changes in diversification correspond to rate increases
in very species-rich genera: \emph{Ypthima} (\texttt{r = 0.311}),
\emph{Charaxes} (\texttt{r = 0.291}), \emph{Callicore} +
\emph{Diaethria} (\texttt{r = 0.220}), \emph{Pedaliodes}
(\texttt{r = 0.196}) and \emph{Taenaris} (\texttt{r = 0.234}). We found
rate increases for other clades as well such as: Mycalesina
(\texttt{r = 0.191}), Oleriina + Ithomiina + Napeogenina + Dircennina +
Godyridina (\texttt{r = 0.187}), Satyrinae (\texttt{r = 0.116}),
Phyciodina in part (\texttt{r = 0.241}) and Satyrina
(\texttt{r = 0.221}), \emph{Coenonympha} (\texttt{r = 0.209}),
\emph{Caeruleuptychia} + \emph{Magneuptychia} (\texttt{r = 0.312}) and
\emph{Taenaris} (\texttt{r = 0.312}). We also found decreases in net
diversification rates for Limenitidinae + Heliconiinae
(\texttt{r = 0.0541}), part of Danaini (\texttt{r = 0.0423}),
Pseudergolinae (\texttt{r = 0.024}) and Coenonymphina
(\texttt{r = 0.065}) (Table 1).

We contrasted these results with those obtained from analyzing the data
in the software BAMM \cite{rabosky20140}. We found that the best model (best
configuration) estimated 2 diversification shifts in our MCC tree
(posterior probability = 0.79) for the Ithomiini and Satyrini (Fig. S1).
Bayes factor (BF) values strongly favored this model in comparison with
a constant rate model (no diversification shifts, BF \(>\) 1,884).

@@

We used MEDUSA to find out whether taking into account the phylogenetic
signal from the random sample of 1000 trees from the posterior
distribution can return similar estimates of diversification to the
values obtained from the MCC tree. We ran MultiMEDUSA on the random
sample of 1000 trees (supp. mat. 5) from the posterior distribution and
compared the results with a MEDUSA analysis on the MCC tree derived from
this sample (supp. mat. 9).

We found that the analysis by MultiMEDUSA on the 1000 trees estimated
lower median net diversification rates for the diversification shifts
found by MEDUSA on the MCC tree derived from the random sample of trees
(Table 2). Although the diversification pattern found by MEDUSA and
MultiMEDUSA was the same, the latter consistently estimated lower rates.
Furthermore, the shifts recovered with low net diversification rate on
the MCC were recovered with negative net diversification rate by
MultiMEDUSA. The background diversification and all shifts found by
MEDUSA on the 1000 trees are provided as an R object in supp. mat. 7.

We also compared the results from MultiMEDUSA (derived from the sample
of 1000 trees) with the splits found by MEDUSA on the MCC tree derived
from this random sample. In the summary statistics, MultiMEDUSA reports
the frequency of the diversification shifts found in the trees
(parameter \texttt{sum.prop}). Thus, if a node is found in only half of
the 1000 trees, but the phylogenetic signal was strong enough to be
picked up by MEDUSA and a node shift was found most of the time, then
the \texttt{sum.prop} should be close to 1. For example, the
\texttt{Charaxes} + \texttt{Polyura} clade was found in only 256 trees,
however MultiMEDUSA was consistently able to find a diversification
shift for that node and the \texttt{sum.prop} value is 0.996.

For the diversification shifts found in both the MCC tree and most of
the samples of 1000 trees (frequency more than 90\%; Table 2), the
MultiMEDUSA approach recovered different rates of diversification than
those found when the MCC tree alone.

There were four diversification shifts found in the trees from the
random sample (Table 2): (i) net diversification rate increase in the
genus \emph{Ypthima} (\(r = 0.22\)); (ii) net diversification rate
increase in the genus \emph{Charaxes} (\(r = 0.21\)); (iii) rate
increase in Ithomiini subtribes Oleriina + Ithomiina + Napeogenina +
Dircennina + Godyridina (\(r = 0.10\)); and (iv) rate increase in
\emph{Callicore} + \emph{Diaethria} (\(r = 0.135\)).

MultiMEDUSA provided mean and standard deviation statistics for the
diversification values on the shifts on the 1000 trees (supp. mat.
13--14), and found that some of the changes in net diversification rate
values had great variation across the posterior distribution of trees. A
boxplot of the net diversification rate values estimated for the clades
that appear in the MCC tree shows that some shifts are estimated as
increased or slowed diversification pace depending on the tree used for
analysis (Fig. 2). This variation is especially wide for the clade
formed by the genera \emph{Magneuptychia} and \emph{Caeruleuptychia}
because MEDUSA estimated diversification values from six times the
background net diversification rate (\texttt{r = 0.5234}) to almost zero
(\texttt{r = 1.22e-01}). The rates for \emph{Taenaris} were between 0.14
and 0.44 (mean value 0.25). Similar degrees of variation were found in
the nodes for \emph{Ypthima}, \emph{Charaxes} and \emph{Coenonympha}
(Fig. 2). The net diversification rates estimates for the clades
(Oleriina + Ithomiina + Napeogenina + Dircennina + Godyridina),
Limenitidinae + Heliconiinae and Pseudergolinae are relatively
consistent across the 1000 trees (Fig. 2).

It is also evident that not all the diversification shifts estimated on
the MCC tree are consistently recovered in most of the 1000 trees. Some
of the splits in the MCC tree are recovered in very few trees, for
example the split for the clade Satyrinae is recovered with a
probability of 0.18 (Fig. 3).

@@

The MEDUSA analyses, taking into account phylogenetic uncertainty,
estimated a net diversification rate increase in part of the clade
Ithomiini across more than 95\% of the trees. Our BiSSE analysis found a
positive effect of the character state ``feeding on Solanaceae'' on the
net net diversification rate on part of Ithomiini (Oleriina + Ithomiina
+ Napeogenina + Dircennina + Godyridina) (Fig. 4). The estimated mean
net diversification rate for taxa that do not feed on Solanaceae was
\texttt{r = 0.11} while the net diversification rate for the Solanaceae
feeders was \texttt{r = 0.16} (see Fig. S2 for a boxplot of speciation
and extinction values for the 95\% credibility intervals). The same
analysis considering \emph{Vanessa} and \emph{Hypanartia} as
non-Solanaceae feeders due to dubious records produced the same pattern
and net diversification rates (Fig. S3). Therefore, the rest of the
analyses were performed assuming these two genera as Solanaceae feeders.
We constrained the BiSSE likelihood model to force equal rates of
speciation for both character states in order to test whether the model
of different speciation rates is a significantly better explanation for
the data. A likelihood ratio test found that the model for increased net
diversification rate for nymphalids feeding on Solanaceae is a
significantly better explanation than this character state having no
effect on diversification (\(\chi^2 = 12.3; 1 df; p < 0.001\)) (Table 3;
character states available in supp. mat. 17, code in supp. mat. 18, and
mcmc run in supp. mat. 19). We combined the post-burnin mcmc generations
from running BiSSE on 250 trees from the random sample of 1000 trees and
found the same pattern as the BiSSE analysis on the maximum clade
credibility tree (combined mcmc run in supp. mat. 20; profiles plot of
speciation rates in Fig. S4; boxplot of 95\% credibility intervals in
Fig. S5). A BiSSE analysis to test whether the trait ``feeding on
Apocynaceae'' had any effect on increased net diversification rates
found similar speciation rates for lineages feeding on Apocynaceae and
other plants (Fig. S6).

@@

\subsubsection*{Ithomiini}

Keith Brown suggested that feeding on Solanaceae was an important event
in the diversification of Ithomiini butterflies \cite{brown1987}. Ithomiini
butterflies are exclusively Neotropical and most species feed on
Solanaceae hostplants during larval stage \cite{willmott2006}. Optimizations of the
evolution of hostplant use on phylogenies evidence a probable shift from
Apocynaceae to Solanaceae in the ancestor of the tribe \cite{willmott2006, brower2006}.
Fordyce (2010) \cite{fordyce2010} found that the Gamma statistics, a LTT plot of
an Ithomiini phylogeny and the fit of the density-dependent model of
diversification are consistent with a burst of diversification in
Ithomiini following the shift from Apocynaceae to Solanaceae.

We investigated whether the strong signal for an increase in net
diversification rate for Ithomiini (found by MEDUSA) can be explained
due to the use of Solanaceae plants as hosts during larval stage. For
this, we used a Bayesian approach \cite{fitzjohn2009} to test whether the trait
``feeding on Solanaceae'' had any effect on the diversification of the
group.
